% Acknowledgements go here
I have many so people to thank for their support during my pursuit of a Ph.D. at the University of Iowa. My advisor, Suely Oliveira, who was always willing to listen to my ideas, even when they weren't very good. I was given many opportunities to be a part of interesting projects and grow as a researcher and leader. Through Dr. Oliveira, I was introduced to Mariana Curi -- I'm appreciative of the time that Dr. Curi spent with me (both in Iowa and virtually), and her willingness to work with me on the research that became my thesis. I also must thank the three other members of my committee, Drs. Jonathan Templin, David Stewart, and Colleen Mitchell for their mentorship, ideas, and guidance.

I had a great opportunity to participate in an internship at ACT, for which I am grateful. In particular, I want to thank Scott and Yuchi for their role in introducing me to new areas of machine learning. The work and guidance provided by them laid the foundation for a second topic to be inserted into my thesis.

I want to thank the Mathematics and Computer Science departments for the challenging coursework and instruction which allowed me to grow as an academic, as well as the Graduate College, all of which provided financial support during my time at Iowa.

Stepping away from math, I also am endebted to my parents, Ken and Karna, for their unwavering encouragement and the abundance of opportunities that they made avaiable to me throughout life. I thank my other family and friends in Iowa City, Des Moines, and elsewhere for their continuous moral support and friendship.

Lastly but certainly not least, I thank my wife Danielle, who has always been available to provide her love, reassurance, and support. She knows me better than anyone, and has stood beside me throughout all the stressful times of graduate school.
